Precise spiking motifs in neurobiological and neuromorphic data

### 2022-10-26 manuscript.tex
Update on Overleaf.

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
Why do neurons communicate through spikes? By definition, a spike, or action potential, is an all-or-none event ---it can occur or not without further details--- at asynchronous timings within a cell assembly, i.e.~it can occur at any continuous time. This stands in stark contrast to the discretized timing classically used in digital processing at the base of modern-day neural networks. In the living world, neural systems almost systematically use this so-called event-based representation. A better understanding of this phenomenon remains a fundamental challenge in neurobiology in order to better interpret the masses of recorded data. It is also an emerging topic in computer science to allow the efficient exploitation of a new class of sensors and event-based computers, called neuromorphic, which could allow significant gains in computing time and energy consumption ---a major societal mission in the age of the digital economy and of global warming.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction: Importance of precise spike timings in the brain}\label{sec:time}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Is there a neural code?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
 Neural activity is directly influenced by our immediate environment and by internal states in order to generate actions. These are essential for survival, which is the sole determinant in the light of natural selection. A central question of modern neuroscience is to better understand the essence of  Neural activity. One sometimes uses the expression ``decoding the neural code'', although this implies the existence of a code, i.e. an explicit representation within the neural activity. Nevertheless, we will use this terminology in all generality. In this respect, it is known that this neural activity is related to a specific variable, and since Galvani's experiments, we know that electrical activity can cause the stretching of a frog's limb (for a review, see~\citep{piccolino_luigi_1997}). A very important characteristic of neural activity is the fact that it consists mainly of action potentials, or spikes, which are brief (about one millisecond) electrochemical impulses that propagate along the axons of neurons. These have the particularity of being essentially binary in their amplitude (i.e. they are prototypical, all or nothing).  The frequency at which these spikes are emitted has been shown to be roughly commensurate with the stretch of the frog's limb~\citep{adrian_impulses_1926}, and this variable (measured in spikes per second, or Hertz) is as of today considered to be the main constituent of the neural code. Neurophysiologists typically use firing sequences to characterize the activity of neurons using different statistics on their individual timing~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.  However, computational neuroscience models have suggested that precise timing may also play a crucial role, and that neurons may be as well integrators than synchrony detectors~\citep{paugam-moisy_computing_2012}. We will investigate this very hypothesis in this review.

Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes that reach its soma. In comparison to a classical analogical vector of inputs, this \emph{event-based representation} present in the neural code is essential in understanding information processing and yet, most neural models do not take advantage of this minute temporal dimension. This hypothesis is directly inspired by neurobiological observations, particularly in the hippocampus, and it expands the capabilities of analog representations (which are based on the firing rate) by considering a representation based on repetitions of spiking motifs at precise times of occurrence. 
Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and also suggest the existence of repeated spatio-temporal spiking motifs in the spike train reaching neurons.  A mathematical formalization would be particularly well suited to neuromorphic computing, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Importantly, validating this hypothesis would also be crucial in our understanding of neuroscientific processes.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Precise timing in the neural code enables efficient vision}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Let us start with a review of the state of the art on the role of dynamics in vision. Vision is the set of processes that allow us to make sense of the luminous world, and is an area intensively studied in neuroscience with respect to the question of neural code. In most mammals, light enters the eye to induce neural activity on the retina, while maintaining a certain similitude between visual space and retinotopy. The origins of this  neuroscientific question can be found in the first experiments of Pierre Flouren showing the relationship between sensations, perception and activity in the cerebral cortex using lesions in animals~\citep{pearce_marie-jean-pierre_2009}. In a series of seminal studies,~\citet{hubel_receptive_1968} showed that this activity could be selective to different features, such as orientation or motion. Remarkably, there is a monotonic relationship between the contrast of visual features and the firing frequency of neurons. However, there is no consensus to explain the multiple nonlinear mechanisms that transform the visual scene into retinotopic neural activity maps, even though these processes seem to constitue essential pieces to this puzzle~\citep{carandini_normalization_2012}.

In particular, there has been some remarkable findings when studying the dynamics of vision. Indeed, Simon Thorpe's group has shown during the last decades numerous examples demonstrating that humans can categorize briefly presented images in a fraction of a second. Its first experiment consisted in asking subjects to categorize images that do or do not contain animals~\citep{thorpe_speed_1996}. The results showed that humans were able to perform this task very well (with a success rate of more than 95\%) but above all that a differential activity for the two categories of images could be observed by electroencephalography, showing that this differentiation emerges at a very early latency in neural activity. These results have been extended to several species including primates but also extended to different experimental protocols and have shown for example that the response could be extremely fast of the order of $120~\ms$ when the task was to perform a saccade~\citep{kirchner_ultra-rapid_2006}. This fast processing also explains the surprising experiments of fast serial detection which consists in presenting a fast succession of different images and to decode via the EEG if the observer can detect for example the presence of an animal. The performances decrease progressively as the frequency of presentation of the images increases. However, it has been shown in the macaque that a significant performance could be maintained with an image presentation time of only $14~\ms$ per image~\citep{keysers_speed_2001}.

This speed of the visual cortex in primates, although surprising, is  compatible with the latencies that are recorded at the neuro-physiological level. Indeed, when an image is presented to the retina, the visual information is rapidly propagated to the thalamus and then to the primary visual cortex takes about $55~\ms$ in the macaque~\citep{nowak_timing_1997}, see Figure~\ref{fig:thorpe}. This functioning of visual processing as a forward pass is most prominent in fast processing but can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}.
%
An important consequence of the speed of processing of vision is that it implies that it is carried out using only very few spikes. Indeed, if we consider that a behavioral response in only $120~\ms$ consists of about ten processing stages following the ``forward'' pathways of the visual system, then this imposes that the processing in a single area is performed with a reduced number of spikes. Altogether, these results show that ``that like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{rucci_temporal_2018}.

%----------------------------%
\begin{figure}
\centering
\note{get the rights or redo a schematics}
\includegraphics[width=.7\textwidth]{figures/visual-latency-estimate.jpg}
\caption{From~\citep{thorpe_seeking_2001} }\label{fig:thorpe}
\end{figure}
%----------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{How timing may encode vectors of real values}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Sensory data is most often represented by continuous values, such as the luminance in the different pixels of a digital image. How may such information be encoded in neuronal activity? The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate is highly irregular~\citep{softky_highly_1993}, which makes it inconsistent with temporal integration of firing rate. %random EPSPs.
% Note also that depending on how an in vitro cell is driven, this may influence the reliability of spike timing in neocortical neurons~\citep{mainen_reliability_1995}.
It has also been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible, where the first spike is aligned but the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{mainen_reliability_1995} (see Figure~\ref{fig:mainen}). When driven by a highly dynamic \emph{frozen} noise, the output spikes are highly reproducible. This is consistent with the differential role of different stimulus frequencies (for instance the gamma range around $80~\si{\Hz}$) on the reliability of spike timing reported in~\citep{nowak_influence_1997} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.'' 
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=.7\textwidth]{figures/MainenSejnowski1995.png} % TODO replace by the simulation
\caption{The timing of the spikes produced following the repetition of a step stimuli is less reproducible than that to a noisy stimulus. While this seems paradoxical at first sight, this is due to the use of the same \emph{frozen} noise at each repetition and highlight the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this \href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995.ipynb}{notebook} for a replication of this result using a simple LIF model.}\label{fig:mainen}
\end{figure}
%----------------------------%

%
At the level of the dynamic processing of visual information, it has been shown in the retina that a coding of luminance values in the image using the timing of the spikes may be at work in the retina~\citep{gollisch_rapid_2008}. In particular, these results show that using  the response of the ganglion cells to the visual gratings projected on the retina could be encoded in the latency of the response and not only in the frequency of the discharge as it is often assumed. These results have been extended to natural images and show a qualitatively similar behavior. The authors' conclusion is that the precise spiking latency of the neurons encodes the spatial features of the image.Interestingly, such a precise latency mechanism may underlie some visual illusions, e.g., the false color illusion in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}. This evidence found in the retina  can be extended to the visual cortex.

@@

In fact, similar results have been demonstrated through neurophysiological recordings in the primary visual cortex and show that different levels of visual activity will induce different levels of neuronal discharge latency in the primary visual area~\citep{celebrini_dynamics_1993}. Many models have used these properties in temporal coding to build fast image categorization networks. These models take the form of artificial spiking neural networks (SNNs) and have been able to demonstrate their practical applications for image categorization~\citep{delorme_ultra-rapid_2000}. First-spike latency codes are a feasible mechanism for information transfer even when biologically plausible estimates of stimulus onset are considered~\citep{chase_first-spike_2007}.  This work has been extended to include unsupervised learning capabilities and we have recently developed a SNN architecture that allows to categorize images of different classes in only a few spikes~\citep{grimaldi_homeostatic_2021,grimaldi_robust_2022}. This type of modeling is extremely important with respect to the development of a new generation of cameras called Silicon Cameras which, instead of using a basic frame-based representation, uses a representation similar to the one we have just described and which consists in representing the image by events~\citep{rasetto_challenges_2022}. This type of modeling often uses the classical architecture of image categorization developed in deep learning while adapting it to the specificity of the event-based representation~\citep{goltz_fast_2021}. Note also that timing is not entirely sensory or internal but can be used as a general coding principle. In~\citep{safaie_turning_2020} for instance, they found that ``timing accuracy was improved when the environment afforded cues that rats can incorporate into motor routines. Timing, at least in animals, may thus be fundamentally embodied and situated.'' 

@@

%----------------------------%
\begin{figure}
\centering
\includegraphics[width=.4\textwidth]{figures/Diesmann_et_al_1999.png}
\caption{Simulation of a synfire propagation \href{https://brian2.readthedocs.io/en/stable/examples/frompapers.Diesmann_et_al_1999.html}{using Brian}. The model consists of 10 groups of 100 neurons each. A pulse with a given jitter is generated in the first group which here generates a pulse after a certain processing delay in the second group with a lower jitter. This allows the propagation of the synchronous activity along the chain of the neural groups.}\label{fig:diesman}
\end{figure}
%----------------------------%
Attempts have been made to detect such synfire chains in neurobiological data~\citep{schrader_detecting_2008}, which envisioned that ``sensitivity is high enough to detect synfire chain activity in simultaneous single-unit recordings of 100 to 200 neurons from such data, enabling application to experimental data in the near future.'' Further models have shown that such synfire chains could be embedded in topographies~\citep{aviel_embedding_2003} while others used conductance-based neurons with feed-forward inhibition to improve the robustness of the propagation~\citep{kremkow_functional_2010}. In particular, this was implemented as a computational neuroscience benchmark model using the pyNN language~\citep{davison_pynn_2008} both in CPU-based and neuromorphic hardware~\citep{pfeil_six_2013} (see Figure~\ref{fig:pynn}).
%
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=.7\textwidth]{figures/fnins-07-00011-g004.jpg}
\caption{A pyNN implementation of a Synfire chain with feedforward inhibition. The background noise  is implemented as random Gaussian current. The state space gives the duration of synfire chains as a function of the protocol's parameters.}\label{fig:pynn}
\end{figure}
%----------------------------%
%
Some experimental results show the emergence of synchrony, for instance in motor cortical function~\citep{riehle_spike_1997}. Interestingly, authors showed that ``Accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.'' Moreover, such emergence could change over the learning period involved in learning a task~\citep{kilavik_long-term_2009} and showed some tuning to movement direction and reaction time~\citep{grammont_spike_2003}. It is important to note that synchronous events tend to lock to LFP beta waves~\citep{denker_lfp_2018}, and was extended to larger assemblies~\citep{torre_synchronous_2016} using statistical methods (see Section~\ref{sec:detection} for further details). Note also that synchronicity may explain some unintuitive results. Indeed it has been shown that thalamocortical synapses are relatively weak compared to the amount of intra-cortical activity. However, this pathway is sufficient to drive the cortex as this input is more often synchronously active~\citep{bruno_cortex_2006}.

@@

%: rhythms as a 
The idea of using the synchrony of coactivation in a cell assembly is an hypothesis that was formalized by~\citet{hebb_organization_1949}. 
This is also expressed in the idea that different cortical areas could achieve binding by synchrony~\citep{fries_mechanism_2005}. The synchronicity will generate rhythms at different range of frequencies with spikes arriving at peak susceptibility (top of a cycle) or down. Such a thory has surprising ly been validated in EEG recordings for example to explain the continuous wagon wheel illusion which corresponds to the pervceived reversal of the rotational movements of the spikes in a rotating wheel~\citep{vanrullen_continuous_2006}. More generallly, it can be shown that the phase of alpha (about $10$ Hz) oscillations  is causally linked with modulations of cortical excitability and with visual perception~\citep{dugue_phase_2011}. 
%
\subsection{A further hypothesis: traveling waves}
To further investigate the role of precise timing, let us also focus on the role of differential timings in an assembly of neurons. As we have seen, a visual feature will induce the firing of different cells at different latencies~\citep{celebrini_dynamics_1993}. Using intra-cellular recordings, it was shown that the response to a focal visual activation would elicit a latency basin, that is a graded onset of the neural response from the most activated to neighboring neurons~\citep{bringuier_horizontal_1999}. In particular, it was shown that the network of horizontal connections within a cortical area is typically not myelinated and that this latency basin would be determined by the propagation speed within the area. For generic visual scenes, this processes would generate a complex interplay between the dynamics of the sensory signal and the spatio- temporal determinants of these interactions and may underly anticipatory mechanisms in the primary visual cortex~\citep{benvenuti_anticipatory_2020,le_bec_horizontal_2022}.
%

Propagating waves occur in many excitable media and were recently found in neural systems from retina to neocortex. While propagating waves are clearly present under anaesthesia, whether they also appear during awake and conscious states remains unclear. One possibility is that these waves are systematically missed in trial-averaged data, due to variability. A recent work~\citep{muller_stimulus-evoked_2014} presents a method for detecting propagating waves in noisy multichannel recordings. Applying this method to single-trial voltage-sensitive dye imaging data, they show that the stimulus-evoked population response in primary visual cortex of the awake monkey propagates as a travelling wave, with consistent dynamics across trials. A network model suggests that this reliability is the hallmark of the horizontal fibre network of superficial cortical layers. Propagating waves with similar properties occur independently in secondary visual cortex, but maintain precise phase relations with the waves in primary visual cortex. These results show that, in response to a visual stimulus, propagating waves are systematically evoked in several visual areas, generating a consistent spatiotemporal frame for further neural interactions.

@@

@@

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
Why do neurons communicate through spikes? By definition, a spike, or action potential, is an all-or-none event which occurs at asynchronous times. In other words, spikes on one side occur or not without further details and second can occur at any continuous time. This stands in stark contrast to the discretized timing classically used in digital processing at the base of modern-day neural networks. In the living world, neural systems almost systematically use this so-called event-based representation. A better understanding of this phenomenon remains a fundamental challenge in neurobiology in order to better interpret the masses of recorded data. It is also an emerging topic in computer science to allow the efficient exploitation of a new class of sensors and event-based computers, called neuromorphic, which could allow significant gains in computing time and energy consumption ---a major societal mission in the age of the digital economy and of global warming.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction: Importance of precise spike timings in the brain}\label{sec:time}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Is there a neural code?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
 Neural activity is directly influenced by our immediate environment and by internal states in order to generate actions. These are essential for survival, which is the sole determinant in the light of natural selection. A central question of modern neuroscience is to better understand the essence of neural activity. One sometimes uses the expression ``decoding the neural code'', although this implies the existence of a code, i.e. an explicit representation within the neural activity. Nevertheless, we will use this terminology in all generality. In this respect, it is known that this neural activity is related to a specific variable, and since Galvani's experiments, we know that electrical activity can cause the stretching of a frog's limb (for a review, see~\citep{piccolino_luigi_1997}). Among various phenomenon, a well studied way of communication between neurons are action potentials, or spikes. In the scope of this article we focus mainly on spikes, which are brief (about one millisecond) electrochemical impulses that propagate along the axons of neurons. These have the particularity of being essentially binary in their amplitude (i.e. they are prototypical, all or nothing).  The frequency at which these spikes are emitted has been shown to be roughly commensurate with the stretch of the frog's limb~\citep{adrian_impulses_1926}, and this variable (measured in spikes per second, or Hertz) is as of today considered to be the main constituent of the neural code. Neurophysiologists typically use firing sequences to characterize the activity of neurons using different statistics on their individual timing~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.  However, computational neuroscience models have suggested that precise timing may also play a crucial role, and that neurons may be as well integrators than synchrony detectors~\citep{paugam-moisy_computing_2012}. We will investigate this very hypothesis in this review.

Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes that reach its soma. In comparison to a classical analogical vector of inputs, this \emph{event-based representation} present in the neural code is essential in understanding information processing and yet, most neural models do not take advantage of this fine-scaled temporal dimension. This hypothesis is directly inspired by neurobiological observations, particularly in the hippocampus, and it expands the capabilities of analog representations (which are based on the firing rate) by considering a representation based on repetitions of spiking motifs at precise times of occurrence. 
Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and also suggest the existence of repeated spatio-temporal spiking motifs in the spike train reaching neurons.  A mathematical formalization would be particularly well suited to neuromorphic computing, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Importantly, validating this hypothesis would also be crucial in our understanding of neuroscientific processes.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Precise timing in the neural code enables efficient vision}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Let us start with a review of the state of the art on the role of dynamics in vision. Vision is the set of processes that allow us to make sense of the luminous world, and is an area intensively studied in neuroscience with respect to the question of neural code. In most mammals, light enters the eye to induce neural activity on the retina, which maintains a certain similitude between external visual space and its representation on the retina. The origins of this  neuroscientific question can be found in the first experiments of Pierre Flouren showing the relationship between sensations, perception and activity in the cerebral cortex using lesions in animals~\citep{pearce_marie-jean-pierre_2009}. In a series of seminal studies,~\citet{hubel_receptive_1968} showed that this activity could be selective to different features, such as orientation or motion. Remarkably, there is a monotonic relationship between the contrast of visual features and the firing frequency of neurons. However, there is no consensus to explain the multiple nonlinear mechanisms that transform the visual scene into retinotopic neural activity maps, even though these processes seem to constitute essential pieces to this puzzle~\citep{carandini_normalization_2012}.

In particular, there have been some remarkable findings when studying the dynamics of vision. Indeed, Simon Thorpe's group has shown during the last decades numerous examples demonstrating that humans can categorize briefly presented images in a fraction of a second. Its first experiment consisted in asking subjects to categorize images that do or do not contain animals~\citep{thorpe_speed_1996}. The results showed that humans were able to perform this task very well (with a success rate of more than 95\%) but above all that a differential activity for the two categories of images could be observed by electroencephalography, showing that this differentiation emerges at a very early latency in neural activity. These results have been extended to several species including primates but also extended to different experimental protocols and have shown for example that the response could be extremely fast of the order of $120~\ms$ when the task was to perform a saccade~\citep{kirchner_ultra-rapid_2006}. This fast processing also explains the surprising experiments of fast serial detection, which consists in presenting a fast succession of different images and to decode via the EEG if the observer can detect for example the presence of an animal~\citep{keysers_speed_2001}. The performances decrease progressively as the frequency of presentation of the images increases. However, it has been shown in the macaque that a significant performance could be maintained with an image presentation time of only $14~\ms$ per image.

This speed of the visual cortex in primates, although surprising, is  compatible with the latencies that are recorded at the neuro-physiological level. Indeed, when an image is presented to the retina, the visual information is rapidly propagated to the thalamus and then to the primary visual cortex, and it takes about $55~\ms$ in the macaque~\citep{nowak_timing_1997}, see Figure~\ref{fig:thorpe}. This functioning of visual processing as a forward pass is most prominent in fast processing but can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}.
%
An important consequence of the speed of processing of vision is that it implies that it is carried out using only very few spikes. Indeed, if we consider that a behavioral response in only $120~\ms$ consists of about ten processing stages following the ``forward'' pathways of the visual system, then this imposes that the processing in a single area is performed with a reduced number of spikes. Altogether, these results show that ``like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{rucci_temporal_2018}.

%----------------------------%
\begin{figure}
\centering
\note{get the rights or redo a schematics}
\includegraphics[width=.7\textwidth]{figures/visual-latency-estimate.jpg}
\caption{From~\citep{thorpe_seeking_2001} }\label{fig:thorpe}
\end{figure}
%----------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{How timing may encode vectors of real values}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Sensory data is most often represented by continuous values, such as the luminance in the different pixels of a digital image. How may such information be encoded in neuronal activity? The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate is highly irregular~\citep{softky_highly_1993}, which makes it inconsistent with temporal integration of firing rate. %random EPSPs.
% Note also that depending on how an in vitro cell is driven, this may influence the reliability of spike timing in neocortical neurons~\citep{mainen_reliability_1995}.
It has also been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible, where the first spike is aligned, but the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{mainen_reliability_1995} (see Figure~\ref{fig:mainen}). When driven by a highly dynamic \emph{frozen} noise, the output spikes are highly reproducible. This is consistent with the differential role of different stimulus frequencies (for instance the gamma range around $80~\si{\Hz}$) on the reliability of spike timing reported in~\citep{nowak_influence_1997} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.'' 
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=.7\textwidth]{figures/MainenSejnowski1995.png} % TODO replace by the simulation
\caption{The timing of the spikes produced following the repetition of a step stimuli is less reproducible than that to a noisy stimulus. While this seems paradoxical at first sight, this is due to the use of the same \emph{frozen} noise at each repetition and highlight the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this \href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995.ipynb}{notebook} for a replication of this result using a simple LIF model.}\label{fig:mainen}
\end{figure}
%----------------------------%

%
At the level of the dynamic processing of visual information, it has been shown in the retina that a coding of luminance values in the image using the timing of the spikes may be at work in the retina~\citep{gollisch_rapid_2008}. In particular, these results show that using  the response of the ganglion cells to the visual gratings projected on the retina could be encoded in the latency of the response and not only in the frequency of the discharge, as it is often assumed. These results have been extended to natural images and show a qualitatively similar behavior. The authors' conclusion is that the precise spiking latency of the neurons encodes the spatial features of the image.Interestingly, such a precise latency mechanism may underlie some visual illusions, e.g., the false color illusion in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}. This evidence found in the retina  can be extended to the visual cortex.

@@

In fact, similar results have been demonstrated through neurophysiological recordings in the primary visual cortex and show that different levels of visual activity will induce different levels of neuronal discharge latency in the primary visual area~\citep{celebrini_dynamics_1993}. Many models have used these properties in temporal coding to build fast image categorization networks. These models take the form of artificial spiking neural networks (SNNs) and have been able to demonstrate their practical applications for image categorization~\citep{delorme_ultra-rapid_2000}. First-spike latency codes are a feasible mechanism for information transfer, even when biologically plausible estimates of stimulus onset are considered~\citep{chase_first-spike_2007}.  This work has been extended to include unsupervised learning capabilities, and we have recently developed a SNN architecture that allows to categorize images of different classes in only a few spikes~\citep{grimaldi_homeostatic_2021,grimaldi_robust_2022}. This type of modeling is extremely important with respect to the development of a new generation of cameras called Silicon Cameras which, instead of using a basic frame-based representation, uses a representation similar to the one we have just described and which consists in representing the image by events~\citep{rasetto_challenges_2022}. This type of modeling often uses the classical architecture of image categorization developed in deep learning, while adapting it to the specificity of the event-based representation~\citep{goltz_fast_2021}. Note also that timing is not entirely sensory or internal, but can be used as a general coding principle. In~\citep{safaie_turning_2020} for instance, they found that ``timing accuracy was improved when the environment afforded cues that rats can incorporate into motor routines. Timing, at least in animals, may thus be fundamentally embodied and situated.'' 

@@

%----------------------------%
\begin{figure}
\centering
\includegraphics[width=.4\textwidth]{figures/Diesmann_et_al_1999.png}
\caption{Simulation of a synfire propagation \href{https://brian2.readthedocs.io/en/stable/examples/frompapers.Diesmann_et_al_1999.html}{using Brian}. The model consists of 10 groups of 100 neurons each. A pulse with a given jitter is generated in the first group which here generates a pulse after a certain processing delay in the second group with a lower jitter. This allows the propagation of the synchronous activity along the chain of the neural groups.}\label{fig:diesman}
\end{figure}
%----------------------------%
Attempts have been made to detect such synfire chains in neurobiological data~\citep{schrader_detecting_2008}, which envisioned that ``sensitivity is high enough to detect synfire chain activity in simultaneous single-unit recordings of 100 to 200 neurons from such data, enabling application to experimental data in the near future.'' Further models have shown that such synfire chains could be embedded in topographies~\citep{aviel_embedding_2003}, while others used conductance-based neurons with feed-forward inhibition to improve the robustness of the propagation~\citep{kremkow_functional_2010}. In particular, this was implemented as a computational neuroscience benchmark model using the pyNN language~\citep{davison_pynn_2008} both in CPU-based and neuromorphic hardware~\citep{pfeil_six_2013} (see Figure~\ref{fig:pynn}).
%
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=.7\textwidth]{figures/fnins-07-00011-g004.jpg}
\caption{A pyNN implementation of a Synfire chain with feedforward inhibition. The background noise  is implemented as random Gaussian current. The state space gives the duration of synfire chains as a function of the protocol's parameters.}\label{fig:pynn}
\end{figure}
%----------------------------%
%
Some experimental results show the emergence of synchrony, for instance in motor cortical function~\citep{riehle_spike_1997}. Interestingly, the authors showed that ``Accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.'' Moreover, such emergence could change over the learning period involved in learning a task~\citep{kilavik_long-term_2009} and showed some tuning to movement direction and reaction time~\citep{grammont_spike_2003}. It is important to note that synchronous events tend to lock to LFP beta waves~\citep{denker_lfp_2018}, and were extended to larger assemblies~\citep{torre_synchronous_2016} using statistical methods (see Section~\ref{sec:detection} for further details). Note also that synchronicity may explain some unintuitive results. Indeed, it has been shown that thalamo-cortical synapses are relatively weak compared to the amount of intra-cortical activity. However, this pathway is sufficient to drive the cortex, as this input is more often synchronously active~\citep{bruno_cortex_2006}.

@@

%: rhythms as a 
The idea of using the synchrony of co-activation in a cell assembly is a hypothesis that was formalized by~\citet{hebb_organization_1949}. 
This is also expressed in the idea that different cortical areas could achieve binding by synchrony~\citep{fries_mechanism_2005}. The synchronicity will generate rhythms at different range of frequencies, with spikes arriving at peak susceptibility (top of a cycle) or down. Such a theory has surprisingly been validated in EEG recordings, for example, to explain the continuous wagon-wheel illusion which corresponds to the perceived reversal of the rotational movements of the spikes in a rotating wheel~\citep{vanrullen_continuous_2006}. More generally, it can be shown that the phase of alpha (about $10$ Hz) oscillations  is causally linked with modulations of cortical excitability and with visual perception~\citep{dugue_phase_2011}. 
%
\subsection{A further hypothesis: traveling waves}
To further investigate the role of precise timing, let us also focus on the role of differential timings in an assembly of neurons. As we have seen, a visual feature will induce the firing of different cells at different latencies~\citep{celebrini_dynamics_1993}. Using intracellular recordings, it was shown that the response to a focal visual activation would elicit a latency basin, that is, a graded onset of the neural response from the most activated to neighboring neurons~\citep{bringuier_horizontal_1999}. In particular, it was shown that the network of horizontal connections within a cortical area is typically not myelinated and that this latency basin would be determined by the propagation speed within the area. For generic visual scenes, these processes would generate a complex interplay between the dynamics of the sensory signal and the spatio-temporal determinants of these interactions and may underlay anticipatory mechanisms in the primary visual cortex~\citep{benvenuti_anticipatory_2020,le_bec_horizontal_2022}.
%

Propagating waves occur in many excitable media and were recently found in neural systems from retina to neocortex. While propagating waves are clearly present under anesthesia, whether they also appear during awake and conscious states remains unclear. One possibility is that these waves are systematically missed in trial-averaged data, due to variability. A recent work~\citep{muller_stimulus-evoked_2014} presents a method for detecting propagating waves in noisy multichannel recordings. Applying this method to single-trial voltage-sensitive dye imaging data, they show that the stimulus-evoked population response in primary visual cortex of the awake monkey propagates as a travelling wave, with consistent dynamics across trials. A network model suggests that this reliability is the hallmark of the horizontal fiber network of superficial cortical layers. Propagating waves with similar properties occur independently in secondary visual cortex, but maintain precise phase relations with the waves in primary visual cortex. These results show that, in response to a visual stimulus, propagating waves are systematically evoked in several visual areas, generating a consistent spatio-temporal frame for further neural interactions.

@@

also

@@

% (mainly based on JOBIM poster)
